---
title: "hotrstuff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hotrstuff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(hotrstuff)
```

This vignette shows how you can download, wrangle, and process ESM data to calculate climate metrics. The idea is that the functions are independent of each other and can still perform their purpose without having to follow the entire workflow detailed below, from start to finish.

This code works for both monthly (`mon`) and daily (`day`) climate data.

I would recommend creating a `data` folder, with the following basic structure:

- `data/raw/`: raw, untouched data
  - `data/raw/wget`: downloaded `wget` scripts
  - `data/raw/tos`: raw ESM data (e.g., tos - sea surface temperature)
- `data/proc/`: processed ESM data
  - `data/proc/merged`: merged files per model - 1 model per model/variable/scenario/frequency combination
  - `data/proc/sliced`: sliced files depending on the required time period
  - `data/proc/yearly`: yearly data (after converting monthly frequency to yearly frequency)
  - `data/proc/regridded`: regridded data, standardizing grids of data
  - `data/proc/ensemble`: ensemble mean/median for each variable/scenario/frequency combination
  
We will follow this folder structure throughout the vignette. It will still work with a different folder structure, just make sure you're inputting the right paths.

# Downloading ESM outputs

To download ESMs, you'd still have to download the `wget` scripts from your chosen CMIP6 repository. I would suggest using: https://aims2.llnl.gov/search.

If you want to follow along, use the wget scripts that are in the `data/raw/wget` folder. But, you can use other `wget` scripts; just make sure to change the arguments.

Then, run the `download_ESM` function to run all the `wget` scripts found in that folder.
```{r download, eval=FALSE}
source(here("R", "functions", "download_ESM.R"))

make_folder(here("data", "raw", "tos")) # make folder
download_ESM(
  indir = here("data", "raw", "wget"), # input directory
  outdir = here("data", "raw", "tos") # output directory
)
```

Un-comment the code above if you want to download new data.

# Merging files

We now merge files by model-variable-scenario-frequency combination.
```{r merge}
source(here("R", "functions", "merge_files.R"))

make_folder(here("data", "proc", "merged", "tos"))
merge_files(
  indir = here("data", "raw", "tos"), # input directory
  outdir = here("data", "proc", "merged", "tos"), # output directory
  year_start = 1985, # earliest year across all the scenarios considered (e.g., historical, ssp126, ssp245, ssp585)
  year_end = 2100 # latest year across all the scenarios considered
)
```

So this would result in just 1 `.nc` file for each model-variable-scenario-frequency combination.

# Adjust and reframe time periods

We now want to only keep the years we're interested in. For example, we want to make sure that for future projections, we only want to look at 2020-2100.

```{r adjust}
source(here("R", "functions", "slice_period.R"))

make_folder(here("data", "proc", "sliced", "tos"))
slice_period(
  indir = here("data", "proc", "merged", "tos"), # input directory
  outdir = here("data", "proc", "sliced", "tos"), # output directory
  frq = "Omon", # ocean, daily
  scenario = "ssp",
  year_start = 2020,
  year_end = 2100,
  overwrite = FALSE
)
```

# Fix calendar periods (if needed)

This function fixes the calendar days, for days with leap years. This is to standardize the calendar across the different models.

```{r calendar}
source(here("R", "functions", "fix_calendar.R"))

fix_calendar(indir = here("data", "proc", "sliced", "tos")) # will be rewritten
```

# Changing frequency of climate data

I created two functions: i) changes frequency to monthly (`monthly_frequency()`), and ii) changes frequency to yearly (`yearly_frequency()`), but because I've started with monthly data here, I didn't run `monthly_frequency()` here (but it works).

```{r monthly}
# source(here("R", "functions", "monthly_frequency.R"))
#
# make_folder(here("data", "proc", "monthly", "tos"))
# monthly_frequency(indir = here("data", "proc", "sliced", "tos"), #input directory
#                  outdir = here("data", "proc", "monthly", "tos"))
```

Uncomment the code above if you're using daily data and changing it to a monthly frequency.

```{r yearly}
source(here("R", "functions", "yearly_frequency.R"))

make_folder(here("data", "proc", "yearly", "tos"))
yearly_frequency(
  indir = here("data", "proc", "sliced", "tos"), # input directory
  outdir = here("data", "proc", "yearly", "tos")
)
```

This should create a new file that has a yearly frequency. For the rest of the analyses, I'll only continue with the yearly data.

# Regridding

The climate models have different grids, so now we need to regrid and standardize the grid. This requires a base, empty raster that will be made automatically in the `regrid_esm()`.

```{r regrid}
source(here("R", "functions", "regrid_esm.R"))
make_folder(here("data", "raw", "base_raster")) # create folder for base raster
make_folder(here("data", "proc", "regridded", "yearly", "tos"))

regrid_esm(
  indir = here("data", "proc", "yearly", "tos"),
  outdir = here("data", "proc", "regridded", "yearly", "tos"),
  blankrast_dir = here("data", "raw", "base_raster"),
  cell_res = 0.25,
  layer = "annual"
)
```

Let's try to plot this. Just plotting for the first time point and for the two models.

```{r plot_regridded}
models <- list.files(here("data", "proc", "regridded", "yearly", "tos"), full.names = TRUE)
model1 <- rast(models[1])
plot(model1$tos_1)
model2 <- rast(models[2])
plot(model2$tos_1)
```

# Create ensemble

For now, we only have 2 models, but if you have more models, you just need to input a list of the model names as a vector in the `model_list` argument.

```{r ensemble}
source(here("R", "functions", "create_ensemble.R"))

make_folder(here("data", "proc", "ensemble", "mean", "tos"))
create_ensemble(
  indir = here("data", "proc", "regridded", "yearly", "tos"), # input directory
  outdir = here("data", "proc", "ensemble", "mean", "tos"), # output directory
  model_list = c("ACCESS-ESM1-5", "CanESM5"), # list of models for ensemble
  variable = "tos", # variable name
  frequency = "Omon", # original frequency of data
  scenario = "ssp126", # emission scenario
  mean = TRUE # if false, takes the median
)
```

Then, we plot to see how it looks.

```{r plot_ensemble}
ensemble_model <- list.files(here("data", "proc", "ensemble", "mean", "tos"), full.names = TRUE)
ensemble <- rast(ensemble_model)
plot(ensemble$tos_1)
```
